Global Distribution of Unintended Pregnancies and Contraceptive Usage

### 2024-04-15 code/analysis-code/statistical-analysis.qmd
Part 5 Updates
Added random forest model and tuned the parameters. Found AIC criteria and analyzed. Added explanation of outcome predictor to clear up confusion.

---
title: "Statistical Analysis"
author: "Taylor Glass and Arlyn Santiago"
date: "03/12/24"
output: html_document
---
## Load necessary packages
```{r}
#| message: false
library(dplyr) # for basic syntax
library(broom) #for cleaning up output from lm()
library(here) #for data loading/saving
library(tidymodels) # for model building 
library(yardstick) # for finding performance metrics 
library(rsample) #for splitting the data into testing/training
library(knitr) # to create performance metric tables 
library(parsnip) # to create a workflow for cross-validation
library(glmnet) # for LASSO regression
```

@@

# save final figure
figure_file = here("results", "figures", "final-figures", "fullmodelpred.png")
ggsave(filename = figure_file, plot=figure1) 
```
The training model predictions, shown in teal, hover pretty closely around the expected relationship between the outcome and the 5 predictors, which is depicted as the 45 degree line. The testing model predictions, shown in coral, are much more extreme than the training model predictions because they are farther from the 45 degree line. While there is some overlap between the two sets of predictions, it is likely that our model is overfitting the data.

@@

We are considering adding a random forest model to further examine the variance in our model.

>>>>

---
title: "Statistical Analysis"
author: "Taylor Glass and Arlyn Santiago"
date: "03/12/24"
output: html_document
---
## Load necessary packages
```{r}
#| message: false
library(dplyr) # for basic syntax
library(broom) #for cleaning up output from lm()
library(here) #for data loading/saving
library(tidymodels) # for model building 
library(yardstick) # for finding performance metrics 
library(rsample) #for splitting the data into testing/training
library(knitr) # to create performance metric tables 
library(parsnip) # to create a workflow for cross-validation
library(glmnet) # for LASSO regression
library(ranger) # for random forest model 
```

@@

# save final figure
figure_file = here("results", "figures", "final-figures", "fullmodelpred.png")
ggsave(filename = figure_file, plot=figure1) 
```
The training model predictions, shown in teal, hover pretty closely around the expected relationship between the outcome and the 5 predictors, which is depicted as the 45 degree line. The testing model predictions, shown in coral, are much more extreme than the training model predictions because they are farther from the 45 degree line. While there is some overlap between the two sets of predictions, it is likely that our model is overfitting the data because the testing observations are so far from the expected relationship.

@@

We also want to explore a random forest model with all 5 main predictors since the previous cross-validation process revealed high overfitting of the multivariate linear regression model. We will use RMSE, MAE, and R squared as performance metrics to compare it to the previous LASSO regression model. 
```{r}
#| message: false
# set seed for reproducibility
set.seed(2468)

# create workflow for random forest model
rf_mod <- rand_forest(mode = "regression") %>% # specify the model type
            set_engine("ranger", seed = 2468) #  set seed for reproducibility internally

random_forest <- rf_mod %>% 
                    fit(pct_upreg_sarc ~ sarc_standard + region + pct_currentlymarried + rate_matdeaths_upreg +
                    rate_matdeaths_abs, data = mydata) # include all predictors

# create predictions using the random forest model
RFpred <- predict(random_forest, new_data = mydata) %>% 
                  select(.pred)

# create a data frame with the predictions and true values 
RFdata <- bind_cols(allpred, mydata$pct_upreg_sarc) %>%
            rename(pct_upreg_sarc = "...2")

# find performance metrics to determine model fit
rmse10 <- rmse(RFdata, truth = pct_upreg_sarc, estimate = .pred)
mae10 <- mae(RFdata, truth = pct_upreg_sarc, estimate = .pred)
rsq10 <- rsq(RFdata, truth = pct_upreg_sarc, estimate = .pred)

# print performance metric results 
print(rmse10)
print(mae10)
print(rsq10)

# visualize random forest predictions verus observed values 
figure4 <-ggplot(RFdata, aes(x=.pred, y= pct_upreg_sarc)) + 
                geom_point() +
                geom_abline(linetype = "dashed") + 
                labs(x="Predicted Values", y="Observed Values", title="Random Forest Model Predictions and Observations")
figure4

# save the random forest figure 
figurefile4 <- here("results", "figures", "final-figures", "randomforestpred.png")
ggsave(filename = figurefile4, plot = figure4)
```
The random forest model has an RMSE of 4.361363, an MAE of 3.386193, and an R-squared of 0.8369754. Compared to the previous model with an RMSE of 4.361363, an MAE of 3.386193, and an R-squared of 0.820600, the random forest model performs very similarly. The RMSE and MAE values are exactly the same beause the same 5 predictors are used for the outcome, but the R-squared value shows that the random forest model performs slightly better. Since a higher R-squared value represents smaller differences between the observed and predicted values, we can conclude that this model is predicting our outcome with slightly higher accuracy. The plot of the predicted versus observed values shows strong correlation for predicted values less than 15, but the deviation from the expected values increases as the predicted values increase.

It is crucial to tune the parameters associated with the random forest model, so we will do that before concluding that it outperforms the LASO regression model. We updated the workflow to allow for tuning of 2 variables: `mtry`, the number of variable to possibly split at in each node, and `min_n`, the minimal node size. We set the number of trees to 300 because it strikes a good balance between achieving the best performance metrics and the processing time required to do so. 
```{r}
#| message: false
# set seed for reproducibility
set.seed(2468)

# update workflow accordingly 
rf_mod2 <- rand_forest(mode = "regression",
                      mtry = tune(),  # allow for tuning
                      min_n = tune(), # allow for tuning
                      trees = 300) %>% 
                          set_engine("ranger", seed = 2468)

# use updated workflow
RFworkflow <- workflow() %>%
  add_recipe(recipe) %>%
  add_model(rf_mod2)

# define a 5 by 5 parameter grid
RFparameter_grid <- grid_regular(mtry(range = c(1,5)),
                            min_n(range = c(1,15)),
                            levels = 5)

# tune the random forest model 
tuned_RFmod <- RFworkflow %>% 
                    tune_grid(resamples = real_samples,
                          grid = RFparameter_grid,
                          metrics = metric_set(rmse))

# observe the tuning process
rf_tune_plot <- tuned_RFmod %>% 
              autoplot()
rf_tune_plot

# save plot 
figure_file5 = here("results", "figures", "final-figures", "randomforesttune.png")
ggsave(filename = figure_file5, plot=rf_tune_plot) 
```
Based on the red line that shows the lowest possible RMSE score, our random forest model performs the best with a node size of 1 and 4 predictors. This begs us to ask the question of which one of the 5 predictors included in our model should be dropped to improve model performance because all previous performance metrics suggested that including all 5 predictors created the best model.

Since both of these models are overfitting the data, we think it is wise to examine an information criterion to determine how well our model might perform on unseen data. Even though cross-validation is a better approach for answering this question because it actually resamples the data, the concern of overfitting justifies finding the AIC value in this case. This measure still evaluates the trade-off between good fit to the data, shown by a low cost function, and model complexity based on the number of parameters. We will compare the single predictor model with the standarized `sarc_standard` variable and the full model because the models have already been shown to have the best performance metrics. 
```{r}
#| message: false
# define single predictor model in format conducive to finding AIC
singlemod <- glm(pct_upreg_sarc ~ sarc_standard, data = mydata)

# find AIC value 
summary(singlemod)

# define full predictor model in format conducive to finding AIC 
fullmod <- glm(pct_upreg_sarc ~ sarc_standard + region + pct_currentlymarried + rate_matdeaths_upreg + rate_matdeaths_abs, data = mydata)

# find AIC value
summary(fullmod)
```

The single predictor model has an AIC of 812.95, and the full predictor model has an AIC of 803.21. The full predictor model has a lower AIC score, and the difference between the two models is 9.74, which means the difference between the two model is meaningful given that it is so close to 10. A lower AIC indicates a better performance, and the rule of thumb is that a difference of 10 between 2 models is meaningful. However, information criterion are not limited by an arbitrary cutoff point, so we feel confident that there is a difference between these two models.

### 2024-04-15 code/eda-code/eda.qmd
Part 5 Updates
Added random forest model and tuned the parameters. Found AIC criteria and analyzed. Added explanation of outcome predictor to clear up confusion.

Plot 3 uses `pct_upreg_sarc` and `sarc_standard` to examine the relationship with short-acting reversible methods. 
```{r}

# explore percentage of unintended pregnancies compared to standardized count of women using short acting reversible methods
plot_3 <- mydata %>% ggplot(aes(x=pct_upreg_sarc, y=sarc_standard)) + 
  geom_point() + 
  geom_smooth(method='lm') + 
  ggtitle("Percent of Unintended Pregnancies vs Standardzied Count of Women Using SARCs") + xlab("Percentage of Unintended Pregnancies") + ylab("Standarized Total Number of Women")
plot(plot_3)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_totalsarc.png")
ggsave(filename = figure_file, plot=plot_3)

>>>>

Plot 3 uses `pct_upreg_sarc` and `sarc_standard` to examine the relationship with short-acting reversible methods. 
```{r}
# explore percentage of unintended pregnancies compared to standardized count of women using short acting reversible methods
plot_3 <- mydata %>% ggplot(aes(x=pct_upreg_sarc, y=sarc_standard)) + 
  geom_point() + 
  geom_smooth(method='lm') + 
  ggtitle("Percent of Unintended Pregnancies vs Standardzied Count of Women Using SARCs") + xlab("Percentage of Unintended Pregnancies") + ylab("Standarized Total Number of Women")
plot(plot_3)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_totalsarc.png")
ggsave(filename = figure_file, plot=plot_3)

### 2024-04-15 products/manuscript/Manuscript.qmd
Part 5 Updates
Added random forest model and tuned the parameters. Found AIC criteria and analyzed. Added explanation of outcome predictor to clear up confusion.

This uses MS Word as output format. [See here](https://quarto.org/docs/output-formats/ms-word.html) for more information.

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(readr)
library(knitr)
library(dplyr)
library(naniar)
library(ggplot2)
```

@@

Step four is to critically analyze the performance of our models to determine which one performs the best in predicting our outcome, which is also done in the statistical-analysis.qmd file in the code > analysis-code folder. We chose 2 performance metrics to find for each of our 9 models: Root Mean Square Error (RMSE) and Mean Absolute Error (MAE). We chose these performance metrics because they are commonly used for linear regression models like the ones we have created. RMSE provides an average magnitude of the error in the predictors, and MAE provides an average absolute difference between the predicted and actual values. We include both metrics to increase the validity of our results. Once we decided on the final model, we use cross-validation to determine how well it will perform on data it has not seen before. This involves splitting the data into testing and training pieces , which allows us to compare the predictions created using both datasets. To examine the uncertainty in the model, we will use the bootstrapping method to resample our data 100 times to find confidence intervals for our predictions. This process is similar to cross-validation because it relies on resamling the training data. Lastly, we used a LASSO regression to add a penalty to each parameter included in the model to determine if any variables should be removed to improve model performance. It is important to tune the hyperparameters associated with LASSO regression to determine which model fits best. We use a grid search to ensure that we try all combinations of the parameters in our grid. We complete the tuning process with a 5-fold cross-validation repeated 5 times. We are still considering adding a random forest procedure to examine the variance in the model. Each of these processes are crucial in determining which model performs best for our dataset. Result tables for each model and test are stored in the results > tables folder. Figures created from cross-validation, boostrapping, and LASSO regression are stored in the results > figure > final-figures folder.

@@

## Statistical analysis
We began with a simple linear regression model with each of the 7 predictor variables to determine the association betwen each variable and the outcome. We explored the region variable further because our main goal is to understand the difference in outcome by region. We used ANOVA to determine the difference between regions, and we added Tukey's Honeslty Significant Difference test to examine pairwise associations. This led to discovering correlations between some of the predictors, which we examined through correlation coefficients. We tried out a combination of these correlated predictors in 2 different models to further examine the interactions. After considering which predictors are most impactful on our outcome, we narrowed the final predictors down to a list of 5. We included a multiple linear regression model with all 5 of these predictors. We used the RMSE and MAE performance metrics to evaluate all 9 models that we created to determine which one fit best. Lastly, we used cross-validation to determine how well our multiple linear regression model performs on data it has not seen before.

@@

The next set of variables we explored were the total counts of women using each birth control method. There were two large outliers that prevented the graphs from being useful. After exploring the outliers, we standardized using the population size of reproductive aged women (15-49 years). There is a positive association between the percentage of unintended pregnancies among short acting reversible method users and the total number of women using that method of birth control, which is to be expected. This association is displayed in @exploratory-result2.
```{r}
#| label: exploratory-result2
#| fig-cap: "Figure 1: Graph Depciting Percent Unintended Pregnancy by Standarized Count of Total Women"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "exploratory-figures", "pctupreg_totalsarc.png"))
```

@@

## Basic statistical analysis
We completed a simple linear regression analysis with each of the 7 predictors shown in @statistical-analysis1. The predictor with the strongest relationship with the outcome is total number of women using short acting reversible methods (`sarc_standard`).

@@

>>>>

```{r, echo=FALSE, message=FALSE}
# load a few R packages for rendering 
library(here)
library(readr)
library(knitr)
library(dplyr)
library(naniar)
library(ggplot2)
```

@@

Step four is to critically analyze the performance of our models to determine which one performs the best in predicting our outcome, which is also done in the statistical-analysis.qmd file in the code > analysis-code folder. We chose 2 performance metrics to find for each of our 9 models: Root Mean Square Error (RMSE) and Mean Absolute Error (MAE). We chose these performance metrics because they are commonly used for linear regression models like the ones we have created. RMSE provides an average magnitude of the error in the predictors, and MAE provides an average absolute difference between the predicted and actual values. We include both metrics to increase the validity of our results. Once we decided on the final model, we use cross-validation to determine how well it will perform on data it has not seen before. This involves splitting the data into testing and training pieces , which allows us to compare the predictions created using both datasets. To examine the uncertainty in the model, we will use the bootstrapping method to resample our data 100 times to find confidence intervals for our predictions. This process is similar to cross-validation because it relies on resamling the training data. Lastly, we used a LASSO regression to add a penalty to each parameter included in the model to determine if any variables should be removed to improve model performance. It is important to tune the hyperparameters associated with LASSO regression to determine which model fits best. We use a grid search to ensure that we try all combinations of the parameters in our grid. We complete the tuning process with a 5-fold cross-validation repeated 5 times. We added a random forest model to attempt to improve model performance. We used another parameter tuning grid to allow for tuning of 2 variables: `mtry`, the number of variable to possibly split at in each node, and `min_n`, the minimal node size. We set the number of trees to 300 because it strikes a good balance between achieving the best performance metrics and the processing time required to do so. Each of these processes are crucial in determining which model performs best for our dataset. Result tables for each model and test are stored in the results > tables folder. Figures created from cross-validation, boostrapping, LASSO regression, and random forest modeling are stored in the results > figure > final-figures folder. Lastly, we examined the AIC scores for the two best performing models since there is strong evidence for overfitting. Even though cross-validation is a better approach for answering the question of how well a model predicted values in unseen data because it actually resamples the data, the concern of overfitting justifies finding the AIC value in this case. This measure still evaluates the trade-off between good fit to the data, shown by a low cost function, and model complexity based on the number of parameters.

@@

## Statistical analysis
We began with a simple linear regression model with each of the 7 predictor variables to determine the association betwen each variable and the outcome. We explored the region variable further because our main goal is to understand the difference in outcome by region. We used ANOVA to determine the difference between regions, and we added Tukey's Honeslty Significant Difference test to examine pairwise associations. This led to discovering correlations between some of the predictors, which we examined through correlation coefficients. We tried out a combination of these correlated predictors in 2 different models to further examine the interactions. After considering which predictors are most impactful on our outcome, we narrowed the final predictors down to a list of 5. We included a multiple linear regression model with all 5 of these predictors. We used the RMSE and MAE performance metrics to evaluate all 9 models that we created to determine which one fit best. Lastly, we used cross-validation to determine how well our multiple linear regression model performs on data it has not seen before. We also used LASSO regression and random forest modeling as forms of subset selection by applying a regularization penalty and generating a tree structure, respectively. Lastly, we examine the AIC values associated with our two best performing models.

@@

The next set of variables we explored were the total counts of women using each birth control method. There were two large outliers that prevented the graphs from being useful. After exploring the outliers, we standardized using the population size of reproductive aged women (15-49 years). There is a positive association between the percentage of unintended pregnancies among short acting reversible method users and the total number of women using that method of birth control, which is to be expected. Our outcome variable is the number of unintended pregnancies among women who are using short-acting reversible contraception, so the number of unintended pregnancies that can be classified in this outcome is dependent on the number of women using this type of contraception. This percentage should increase as the number of women using short acting-reversible contraception increases because more women are at risk of experiencing an unintended pregnancy due to this contraception if they are using this type of contraception in the first place. This outcome variable does not speak to the effectiveness of short-acting reversible contraception, so it makes sense that our outcome would increase as more women put themselves at risk of experiencing the outcome through becoming classified as short-acting reversible contraception users. This association is displayed in @exploratory-result2.
```{r}
#| label: exploratory-result2
#| fig-cap: "Figure 1: Graph Depciting Percent Unintended Pregnancy by Standarized Count of Total Women"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "exploratory-figures", "pctupreg_totalsarc.png"))
```

@@

## Basic statistical analysis
We completed a simple linear regression analysis with each of the 7 predictors shown in @statistical-analysis1. The predictor with the strongest positive relationship with the outcome is total number of women using short acting reversible methods (`sarc_standard`). This percentage should increase as the number of women using short acting-reversible contraception increases because more women are at risk of experiencing an unintended pregnancy due to this contraception if they are using this type of contraception in the first place.

@@

After creating the random forest model, we found that the RMSE and MAE values are exactly the same because the same 5 predictors are used for the outcome, but the R-squared value shows that the random forest model performs slightly better. Since a higher R-squared value represents smaller differences between the observed and predicted values, we can conclude that this model is predicting our outcome with slightly higher accuracy. The plot of the predicted versus observed values shown in @statistical-analysis11 displays stronger correlation for predicted values compared to the testing and training done with cross-validation previously.  
```{r}
#| label: statistical-analysis11
#| fig-cap: "Predicted vs Observed Values in Random Forest Model"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "final-figures", "randomforestpred.png"))
```

Again, it is crucial to tune hyperparameters associate with machine learning models. Based on the red line displayed in @statistical-analysis12 that shows the lowest possible RMSE score, our random forest model performs the best with a node size of 1 and 4 predictors. This begs the question of which one of the 5 predictors included in our model should be dropped to improve model performance because all previous performance metrics suggested that including all 5 predictors created the best model. However, there is a minimal increase in RMSE when all five models are included in the model with a node size of 1.  
```{r}
#| label: statistical-analysis12
#| fig-cap: "Tuning Parameters for Random Forest Model"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "final-figures", "randomforesttune.png"))
```
